---
title: "Simulations Pre-Class Project"
date: "Due March 13, 2017 at 5:00pm"
output:
  html_document


---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>
```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```




#Project Goals:


With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
    a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 
```{r}
## open libraries 
library(tidyverse)
## set seed for replicability, running the simulation 
set.seed(1234)
## function for gambler's ruin 
gamb.ruin <- function(bet, bankroll, p) {
  money.back <- bet
  x <- 0 #empty vector to count number of games 
  while(bankroll >= bet) { ##run as long as you have money to bet
    game <- runif(1)
    if(game <= p) { ## if win, bet goes back to original value 
      bankroll = bankroll + bet
      bet = money.back
    } else { ## if lose, next bet doubles, want money back wants to get money back
      bankroll = bankroll - bet
      bet = bet + bet 
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
  }
  times.played <- x 
  return(times.played)
}

num.games <- numeric() ##create empty numeric vector 
simulation <- 5000
for(i in 1:simulation) {
  hand <- gamb.ruin(bet=100, bankroll=900, p=.5) ##already betting 100, so you only have 900 to really play, place number of games in vector 
  num.games <- c(num.games, hand)
}

a <- length(num.games) #simualted games 
b <- length(
  which(num.games<=100) #amount of simulations that ended at/before 100 hands
)
a
b

PR <- b/a
PR
```
    b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
```{r}
a <- length(num.games)
b <- length(
  which(num.games <=500)
)

PR <- b/a
PR
```
    c. the mean time you go bust, given that you go bust within the first 5000 hands.
```{r}
mean(
  num.games[which(num.games<=5000)]
)
```
    d. the mean and variance of your bankroll after 100 hands (including busts).
```{r}
## set seed for replicability, running the simulation 
set.seed(1234)
## function for gambler's ruin but w/ 100 games
gamb.ruin.bankroll <- function(bet, bankroll, p) {
  x <- 0
  money.back <- bet 
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {  
      bankroll = bankroll + bet
      bet = money.back
    } else { 
      bankroll = bankroll - bet
      bet = bet + bet
    }
    if(bankroll < bet){break} 
    if(x == 100){break} ##end if hit 100 games 
  }
  x <- x+1
  times.played <- x 
  return(bankroll) ## return bankroll instead of num of games
}

bankroll <- numeric() ## set up empty bankroll 
simulation <- 5000
for(i in 1:simulation){
  hand <- gamb.ruin.bankroll(100, 900, 0.5)
  bankroll <- c(bankroll, hand)
}

mean(bankroll)
var(bankroll)

```
    e. the mean and variance of your bankroll after 500 hands (including busts).
```{r}
## set seed for replicability, running the simulation 
set.seed(1234)
## function for gambler's ruin, but now with 500 games
gamb.ruin.bankroll.500 <- function(bet, bankroll, p) {
  x <- 0 
  money.back <- bet
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {
      bankroll = bankroll + bet
      bet = money.back
    } else { 
      bankroll = bankroll - bet
      bet = bet + bet
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet
    if(x == 500){break} ##end if hit 500 games 
  }
  times.played <- x 
  return(bankroll)
}

bankroll <- numeric() ## set up empty bankroll 
simulation <- 5000
for(i in 1:simulation){
  hand <- gamb.ruin.bankroll.500(100, 900, 0.5)
  bankroll <- c(bankroll, hand)
}

mean(bankroll)
var(bankroll)
```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
    a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.
```{r}
## set seed for replicability, running the simulation 
set.seed(1234)
## basically gamb.ruin function with edits to p 
m.chains <- function(bet, bankroll, p) {
  x <- 0 
  chances <- p #create a vector for chance at winning at original p 
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) { ## bets remain constant, p change 
      bankroll = bankroll + bet
      p = p + 0.01
    } else { ## if lose, p goes back to original 
      bankroll = bankroll - bet
      p = chances
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
    if(x == 100000){break}
  }
  times.played <- x
  return(bankroll)
}

collect <- numeric()
simulation <- 99
for(i in 1:simulation){
  hand <- m.chains(100, 900, 0.48)
  collect <- c(collect, hand)
}
mean(collect)
## game is not fair because mean of your returns are 0... 
```
  
  b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?
```{r}
set.seed(1234)
return <- numeric()
for (i in 1:15) { ## create a loop that runs m.chains but changes the beginning p everytime 
  return.size <- numeric() #empty vector for inputs from m.chains
  simulated <- 100
    for(x in 1:simulated) { 
      hand <- m.chains(100, 900, 0.48 + (0.002*i))
      return.size <- c(return.size, hand)
    }
  prop.return.0 <- length(
    which(return.size == 0)
  )/100 ## proportion of returns == 0, should be close to 50%
  return <- c(return, prop.return.0)
}
return

#when starting probability is 51% the game is most fair, slightly in favor of the gambler @ .49 
```
  c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 
```{r}
set.seed(1234)
## basically m.chains but with new arguement 
m.chains.increase <- function(bet, bankroll, p, p.increase) {
  x <- 0 
  chances <- p  
  while(bankroll >= bet) { 
    game <- runif(1)
    if(game <= p) {
      bankroll = bankroll + bet
      p = p + p.increase
    } else { ## if lose, p goes back to original 
      bankroll = bankroll - bet
      p = chances
    }
  x <- x+1
    if(bankroll < bet){break} ##end if you don't have enough to bet 
    if(x == 100000){break}
  }
  times.played <- x
  return(bankroll)
}

## now to test things via loops
return <- numeric()
for (i in 1:15) { 
  return.size <- numeric() 
  simulated <- 100
    for(x in 1:simulated) { ## change code so i  now affects increment 
      hand <- m.chains.increase(100, 900, 0.48, .003*i)
      return.size <- c(return.size, hand)
    }
  prop.return.0 <- length(
    which(return.size == 0)
  )/100 ## proportion of returns == 0 
  return <- c(return, prop.return.0)
}
return
#somewhere between increments of .30 (56%) and .33 (46%) it is the most fair
```
